\chapter{About this course}

\section{Introduction}
This course is an introduction into the wonderful world of Android.
The aim of this course is not to turn you into a specialized Android developer, but to give an introduction to the general concepts of the Android framework and provide the tools you need to find your way in this fragmented world. 

Starting from first semester of 2018 we will adopt Kotlin as the default language for developing Android applications. 

\section{ECTS}
ECTS is an acronym for European Credit Transfer and Accumulation System.
This stipulates every final attainment level you should have when finishing this course.
These can be found \href{https://bamaflexweb.hogent.be/BMFUIDetailxOLOD.aspx?a=98522\&b=5\&c=1}{here}. 

\section{Course Material}
The course is taught using the following materials:

\begin{itemize}
	\item These course notes
	\item Book \textit{Kotlin for Android Developers} \cite{Leiva2018}
	\item Book  \textit{The Busy Coder's Guide to Android Development} \cite{murphymarkl.2017}
	\item External resources and references (these course notes, Chamilo links \dots)
	\item The usefull post from ProdAndroidDev \url{https://proandroiddev.com/}
	\item TODO: boeken toevoegen die we gebruikt hebben
\end{itemize}

\section{Course organisation}
This course consists of 12 lessons.
The first part of each lesson will cover the theoretical aspects and the second part will put theory into practice.
Normally you will not be able to complete the exercises in one hour so it is up to you to complete these exercises at home.
If you encounter any difficulties  you are welcome to ask questions in the next lesson or via the forum on Chamilo.
We don't encourage sending personal e-mail: you are not alone in having difficulties with certain topics.
E-mail is not the best way for us to help you.

Think carefully when you write down your question: make it reproducible. Follow these guidelines: \href{https://stackoverflow.com/help/how-to-ask}{https://stackoverflow.com/help/how-to-ask}

\subsection{Covered Topics}
This course will roughly cover the following topics:

\begin{enumerate}
	\item History of  Android and Intro Kotlin
	\item Activities + layout files 
	\item Structure of an  app, Activity lifecycle, Resources, buildproces
	\item Fragments + UI components and layouts 
	\item Intents, BroadcastReceivers (EventBus) and Parcable 
	\item recyclerview + cardview 
	\item Android coding practices  + common Android software architectures  
	\item Persistentence in Android
	\item Networking
	\item Testing in Android
	\item Composer for android testing
	\item Memory leaks 
	\item Firebase 
\end{enumerate}

\section{Prior knowledge}
This course assumes knowledge of the Java programming language.
You do not need to know everything, but you need to understand the following concepts:

\begin{itemize}
	\item Language fundamentals (flow control, etc.)
	\item Classes and objects
	\item Methods and data members
	\item Public, private, and protected
	\item Static and instance scope
	\item Exceptions
	\item Threads
	\item Collections
	\item Generics
	\item File I/O
	\item Reflection
	\item Interfaces
\end{itemize}

\section{Examination}
Both for the first as for the second exam period, you will be examined via an oral exam.  The first part is a theory part, where 
you will be handed a set of questions and  you will have time to prepare them. You will be given the time to defend your answer.

For the second part, you will need to create your own application which you will need to present on the examination. You do not need to prepare a powerpoint, but you will be asked some questions regarding your own application mainly on how you have tackled some of the technical issues. Your application must at least have the following requirements.

\begin{enumerate}
	\item An extensive presentation layer obeying the rules defined by Android (i.e. the Android style guide) and making use of all the best practices presented in these course notes.
	\item An application layer obeying the rules defined by Android and making use of all the design patterns presented in these course notes.
	\item A domain layer
	\item A data access layer
	\begin{enumerate}
		\item A persitence layer able to store complex data in a structure way
		\item A networking layer able to make requests, download data \dots
		\item A testing layer, consisting severale carefully considered tests. Trivial tests are not sufficient. 
		\item Other service layers needed to support the aforementioned layers. 
	\end{enumerate}
\end{enumerate}

Your application must:
\begin{itemize}
	\item Scale well and adapt to the screen where it is used (at least mobile phone and tablet)
	\item Have a UI which follows basic principle UI and UX rules.
	\item Make use of carefully considered libraries
	\item Make use of a carefully considered package structured.
	\item Make use of all the best practices defined by Android.
	\item Be well documented (commentary) and have a carefully considered repository documentation.
	\item Should be developed using all best practices of git.
\end{itemize}



